Amplicon sequence variants artificially split bacterial genomes into separate clusters

### 2021-02-26 submission/manuscript.Rmd
Fifth round of revisions

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
editor_options:
  chunk_output_type: console
bibliography: references.bib
csl: asm.csl
---

@@

format_pct <- function(x, my_digits=2){

@@

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017]. Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013]. Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters. Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA gene per genome and the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014]. Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes. The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*). As **rrn** copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome. However, across those accessions, there were `r single_nasvs` versions of the gene. An *E. coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median of `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`). Across the `r ecoli_copy$N` *E. coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` versions of the gene. These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their distances between each other. Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters. To control for uneven representation of genomes across species, I randomly selected one genome from each species for the following analyses and repeated each randomization 100 times. I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1). Among species with 7 copies of the *rrn* operon (e.g., *E. coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 2). Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively. But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2). Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters. At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if using ASVs or too fine of a threshold to define OTUs. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019]. Since there is no consensus for a biologicaly definition of a bacterial species [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how bacterial species are named is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]). This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005]. Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition. A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014]. There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021]. A short sequence from a bacterial genome simply cannot differentiate between species. Moreover, it is difficult to defend a clustering approach that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possibility that parts of a genome would have different ecologies. Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.

@@

**(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur. The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017]. Distance thresholds between 0.25 and 10.0% in 0.25 percentage point increments were used to assign sequences to OTUs.

**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species, I randomly selected one genome from each species. The random selection was repeated 100 times. Analyses based on this randomization reported the median of the 100 randomizations. The intraquartile range between randomizations was less than `r lump_split_iqr`. Because the range was so small, the confidence intervals were more narrow than the thickness of the lines in Figures 1 and 2 and were not included.

@@

**(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences. Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19). For this reason, when an ASV splits a genome into different units, those units were called clusters rather than OTUs.

@@

**Figure 2. As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of ASVs and OTUs representing multiple species increased.** These data represent the median fractions for both measurements across 100 randomizations. In each randomization, one genome was sampled from each species.

>>>>

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \raggedright
editor_options:
  chunk_output_type: console
bibliography: references.bib
csl: asm.csl
---

@@

format_pct <- function(x, my_digits=1){

@@

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017]. Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013]. Yet, ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters. Conversely, using too broad of a threshold to define OTUs could cluster together multiple bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA gene per genome and the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014]. Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes. The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*). As *rrn* copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome. However, across those accessions, there were `r single_nasvs` versions of the gene. An *E. coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median of `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`). Across the `r ecoli_copy$N` *E. coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` versions of the gene. These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their distances between each other. Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters. To control for uneven representation of genomes across species, I randomly selected one genome from each species and repeated each randomization 100 times. I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1). Among species with 7 copies of the *rrn* operon (e.g., *E. coli*), a distance threshold of `r format_pct(threshold_v19, 2)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 1). Similarly, thresholds of `r format_pct(threshold_v34, 2)`, `r format_pct(threshold_v4, 2)`, and `r format_pct(threshold_v45, 2)`% were required for the V3-V4, V4, and V4-V5 regions, respectively. But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 1). Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.

At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (Figure 2). For this analysis, I again randomly selected one genome from each species to control for uneven representation of genomes across species and repeated each randomization 100 times. Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if using ASVs or too fine of a threshold to define OTUs. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019]. Since there is no consensus for a biologicaly definition of a bacterial species [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how bacterial species are named is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]). This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005]. Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition. A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014]. There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021]. A short sequence from a bacterial genome simply cannot differentiate between species. Moreover, it is difficult to defend a clustering threshold that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possibility that different *rrn* operons from the same genome would have different ecologies. Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.

@@

**(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur. The OptiClust algorithm, as implemented in mothur, was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017]. Distance thresholds between 0.25 and 10.00% in 0.25 percentage point increments were used to assign sequences to OTUs.

**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species (Figures 1 and 2), I randomly selected one genome from each species. The random selection was repeated 100 times. Analyses based on this randomization reported the median of the 100 randomizations. The intraquartile range between randomizations was less than `r lump_split_iqr`. Because the range was so small, the confidence intervals were more narrow than the thickness of the lines in Figures 1 and 2 and were not included.

@@

**(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences. Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-26). For this reason, when an ASV split a genome into different units, those units were called clusters rather than OTUs.

@@

**Figure 2. As the distance threshold used to define an OTU increased, the percentage of ASVs and OTUs representing multiple species increased.** These data represent the median fractions for both measurements across 100 randomizations. In each randomization, one genome was sampled from each species.
